---
title: "Summary of Results"
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Marie-Abèle Bind 
    url: https://scholar.harvard.edu/marie-abele
    affiliation: Biostatistics Center, Massachusetts General Hospital
    affiliation_url: https://biostatistics.massgeneral.org/faculty/marie-abele-bind-phd/
  - name: Léo Zabrocki 
    url: https://lzabrocki.github.io/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/fr/zabrocki-leo/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  fig.path = "images/",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "CairoPNG",
  dpi = 400
)
```

<style>
body {
text-align: justify}
</style>

In this document, we explain why it is very important to check covariates balance and the common support of the data in observational studies. We illustrate these two issues by adapting an example from [Jennifer Hill (2011)](https://www.tandfonline.com/doi/abs/10.1198/jcgs.2010.08162). **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**.

# Required Packages and Data Loading

To reproduce exactly the `intuition.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `intuition.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(Cairo) # for printing custom police of graphs
```

We load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(here::here("inputs", "2.functions",
                  "script_theme_tufte.R"))
# define nice colors
my_blue <- "#0081a7"
my_orange <- "#fb8500"
```


# Fake-Data Simulation

We simulate below a fake dataset of 


2000 daily observations where

```{r}
# set seed
set.seed(42)

# set sample size
sample_size <- 2000

data <- tibble(heat_wave = rbinom(sample_size, size = 1, p = 0.1)) %>%
  rowwise() %>%
  mutate(
    o3_lag = ifelse(heat_wave == 1, rnorm(1, mean = 50, sd = 10), rnorm(1, mean = 20, sd = 10)) %>% abs(.),
    y_0 = rnorm(1,  1500 + 200*log(o3_lag), 250),
    y_1 = rnorm(1, 2300 + exp(o3_lag*0.11), 250),
    y_obs = ifelse(heat_wave==1, y_1, y_0)
  ) %>%
  ungroup()
```


```{r, fig.width=6, fig.height=4, code_folding="Please show me the code!"}
data %>%
  ggplot(., aes(x = o3_lag, y = y_0)) +
  geom_point(shape = 16, colour = my_blue, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = my_orange) +
  labs(x = "O3 in t-1 (µg/m³)", y = "Y(0)", title = "Y(0) vs. Ozone Concentration in t-1") +
  theme_tufte()
```


```{r, fig.width=6, fig.height=4, code_folding="Please show me the code!"}
data %>%
  mutate(heat_wave = ifelse(heat_wave==1, "With Heat Wave", "No Heat Wave")) %>%
  ggplot(., aes(x = o3_lag, y = y_obs, colour = heat_wave)) +
  geom_point(shape = 16, alpha = 0.5) +
  scale_colour_manual(name = "Group:", values = c(my_blue, my_orange)) +
  labs(x = "O3 in t-1 (µg/m³)", y = "YLL", title = "Y(0) vs. Ozone Concentration in t-1") +
  theme_tufte()
```

```{r, fig.width=6, fig.height=4, code_folding="Please show me the code!"}
data %>%
  pivot_longer(cols = c(y_0, y_1), names_to = "potential_outcome", values_to = "value") %>%
  ggplot(., aes(x = o3_lag, y = value, colour = potential_outcome)) +
  geom_point(shape = 16, alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
  scale_colour_manual(name = "Group:", values = c(my_blue, my_orange)) +
  labs(x = "O3 in t-1 (µg/m³)", y = "YLL", title = "Y(0) vs. Ozone Concentration in t-1") +
  theme_tufte()
```

```{r}
data %>%
  filter(heat_wave==1) %>%
  summarise(mean(y_1-y_0))
```

```{r}
data %>%
  summarise(mean(y_1-y_0))
```

```{r}
lm(y_obs ~ heat_wave + o3_lag, data = data) %>%
  summary()
```



```{r}
library(MatchIt)
# match without caliper
matching_ps_0.5_caliper <-
  matchit(
    heat_wave ~ o3_lag,
    caliper = 0.1,
    data = data
  )

# display summary of the procedure
matching_ps_0.5_caliper

# we retrieve the matched data
data_ps_0.5_caliper <- match.data(matching_ps_0.5_caliper)

# compute the true effect for the data
round(mean(data_ps_0.5_caliper$y_1[data_ps_0.5_caliper$heat_wave == 1] - data_ps_0.5_caliper$y_0[data_ps_0.5_caliper$heat_wave == 0]), 0)
round(mean(data_ps_0.5_caliper$y_1[data_ps_0.5_caliper$heat_wave == 1] - data_ps_0.5_caliper$y_0[data_ps_0.5_caliper$heat_wave == 1]), 0)
```


















