---
title: "Summary of Results"
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Marie-Abèle Bind 
    url: https://scholar.harvard.edu/marie-abele
    affiliation: Biostatistics Center, Massachusetts General Hospital
    affiliation_url: https://biostatistics.massgeneral.org/faculty/marie-abele-bind-phd/
  - name: Léo Zabrocki 
    url: https://lzabrocki.github.io/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/fr/zabrocki-leo/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  fig.path = "images/",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "CairoPNG",
  dpi = 400
)
```

<style>
body {
text-align: justify}
</style>

In this document, we display the results of the three matching procedures we implemented. **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**.

# Required Packages and Data Loading

To reproduce exactly the `summary_results.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `summary_results.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(Cairo) # for printing custom police of graphs
```

We load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(here::here("inputs", "2.functions",
                  "script_theme_tufte.R"))
# define nice colors
my_blue <- "#0081a7"
my_orange <- "#fb8500"
```

# Covariates Balance Results

We load and clean the data on covariates balance for each matching procedure:

```{r}
# load and bind data
files <- dir(
  path = here::here("inputs",
                    "3.outputs",
                    "1.data",
                    "covariates_balance"),
  pattern = "*.RDS",
  full.names = TRUE
)

data_cov_balance <- files %>%
  map( ~ readRDS(.)) %>%
  reduce(rbind)

# recode type
data_cov_balance <- data_cov_balance %>%
  mutate(
    type = case_when(
      type == "Binary" ~ "Binary",
      type == "Contin." ~ "Continuous",
      type == "binary" ~ "Binary",
      type == "continuous" ~ "Continuous"
    )
  )

# clean data for summary statistics
data_cm_cpm <- data_cov_balance %>%
  filter(matching_procedure %in% c("Coarsened Exact Matching", "Constrained Pair Matching")) %>%
  filter(sample == "Matched Data") %>%
  select(var, type, stat, matching_procedure)

data_initial <- data_cov_balance %>%
  filter(matching_procedure == "Coarsened Exact Matching" &
           sample == "Initial Data") %>%
  select(-matching_procedure) %>%
  rename(matching_procedure = sample)

data_ps <- data_cov_balance %>%
  filter(sample %in% c("Without a Caliper", "With a 0.5 SD Caliper")) %>%
  mutate(
    matching_procedure = ifelse(
      sample == "Without a Caliper",
      "Propensity Score Matching without a Caliper",
      "Propensity Score Matching with a 0.5 SD Caliper"
    )) %>%
  select(-sample)

# bind clean data
data_cov_balance <- bind_rows(data_initial, data_ps) %>%
  bind_rows(., data_cm_cpm)
```

We compute summary statistics on balance for continuous variables:

```{r, echo = FALSE, layout="l-body-outset"}
data_cov_balance %>%
  filter(type == "Continuous") %>%
  group_by(matching_procedure) %>%
  summarise(
    "Average of Standardized Mean Differences" = round(mean(stat), 2),
    "Std. Deviation" = round(sd(stat), 2)
  ) %>%
  ungroup() %>%
  mutate(
    matching_procedure = fct_relevel(
      matching_procedure,
      "Initial Data",
      "Propensity Score Matching without a Caliper",
      "Propensity Score Matching with a 0.5 SD Caliper",
      "Coarsened Exact Matching",
      "Constrained Pair Matching"
    )
  ) %>%
  arrange(matching_procedure) %>%
  rename("Matching Procedure" = matching_procedure) %>%
  kable(., align = c("l", "c", "c"))
```

We compute summary statistics on balance for binary variables:

```{r, echo = FALSE, layout="l-body-outset"}
data_cov_balance %>%
  filter(type == "Binary") %>%
  group_by(matching_procedure) %>%
  summarise(
    "Average of Proportion Differences" = round(mean(stat), 2),
    "Std. Deviation" = round(sd(stat), 2)
  ) %>%
  ungroup() %>%
  mutate(
    matching_procedure = fct_relevel(
      matching_procedure,
      "Initial Data",
      "Propensity Score Matching without a Caliper",
      "Propensity Score Matching with a 0.5 SD Caliper",
      "Coarsened Exact Matching",
      "Constrained Pair Matching"
    )
  ) %>%
  arrange(matching_procedure) %>%
  rename("Matching Procedure" = matching_procedure) %>%
  kable(., align = c("l", "c", "c"))
```

# Analysis Results

We load the results on the estimates of treatment effect sizes for the matching procedures and the regression outcome approach:

```{r}
# load and bind data
files <- dir(
  path = here::here("inputs",
                    "3.outputs",
                    "1.data",
                    "analysis_results"),
  pattern = "*.RDS",
  full.names = TRUE
)

data_analysis_results <- files %>%
  map(~ readRDS(.)) %>% 
  reduce(rbind)  
```

We display below the summary of results in a table:

```{r, echo = FALSE, layout="l-body-outset"}
data_analysis_results %>%
  mutate("Width of CI" = conf.high - conf.low ,
         "95% CI" = paste(paste0("(", conf.low, ";"),  paste0(conf.high, ")"))) %>%
  select(procedure, sample_size, estimate, `95% CI`, `Width of CI`) %>%
  rename("Procedure" = procedure,
         "Sample Size" = sample_size,
         "Estimate" = estimate) %>%
  mutate(
    Procedure = fct_relevel(
      Procedure,
      "Outcome Regression Model without Covariates Adjustment",
      "Outcome Regression Model with Covariates Adjustment",
      "Propensity Score without a Caliper",
      "Propensity Score without a Caliper and with Covariates Adjustment",
      "Propensity Score with a 0.5 SD Caliper",
      "Propensity Score with a 0.5 SD Caliper and with Covariates Adjustment",
      "Coarsened Matching without Covariates Adjustment",
      "Coarsened Matching with Covariates Adjustment",
      "Constrained Pair Matching without Covariates Adjustment",
      "Constrained Pair Matching with Covariates Adjustment"
    )
  ) %>%
  arrange(Procedure) %>%
  kable(., align = c("l", "c", "c", "c", "c"))
```

We display below the summary of results in a graph:

```{r, echo = FALSE, fig.width = 10, fig.height = 4, layout="l-body-outset", code_folding="Please show me the code!"}
# add column for stripes
data_analysis_results <- data_analysis_results %>%
  mutate(group_index = rep(1:5, each = 2), 
         stripe = ifelse((group_index %% 2) == 0, "Grey", "White"))

# make the graph
graph_results_ci <-
  data_analysis_results %>%
    mutate(
    procedure = fct_relevel(
      procedure,
      "Outcome Regression Model without Covariates Adjustment",
      "Outcome Regression Model with Covariates Adjustment",
      "Propensity Score without a Caliper",
      "Propensity Score without a Caliper and with Covariates Adjustment",
      "Propensity Score with a 0.5 SD Caliper",
      "Propensity Score with a 0.5 SD Caliper and with Covariates Adjustment",
      "Coarsened Matching without Covariates Adjustment",
      "Coarsened Matching with Covariates Adjustment",
      "Constrained Pair Matching without Covariates Adjustment",
      "Constrained Pair Matching with Covariates Adjustment"
    )
  ) %>%
  ggplot(.,
         aes(
           x = estimate,
           y = fct_rev(procedure),
         )) +
    geom_rect(
    aes(fill = stripe),
    ymin = as.numeric(as.factor(data_analysis_results$procedure))-0.495,
    ymax = as.numeric(as.factor(data_analysis_results$procedure))+0.495,
    xmin = -Inf,
    xmax = Inf,
    color = NA,
    alpha = 0.4
  ) +
  geom_vline(xintercept = 0) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.5, colour = my_blue, lwd = 0.8) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_fill_manual(values = c('gray86', "white")) +
  guides(fill = FALSE) +
  ylab("") + xlab("Estimates for Heat Waves Effects on YLL") +
  theme_tufte() +
  theme(axis.ticks.y = element_blank())


# print the graph
graph_results_ci

# save the graph
ggsave(
  graph_results_ci,
  filename = here::here(
    "inputs",
    "3.outputs",
    "2.graphs",
    "graph_results_ci.pdf"
  ),
  width = 25,
  height = 12,
  units = "cm",
  device = cairo_pdf
)
```

